The Perils of Exploration under Competition: \\ A Computational Modeling Approach

### 2020-09-24 journal paper/ec19paper/content/inverted_u.tex
change around spacing
Adjusted placement of figures and tables a bit.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\begin{table*}[t]
\centering
\begin{adjustbox}{width=\textwidth,center}
\begin{tabular}{|c|c|c|c||c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail}
  & \multicolumn{3}{c|}{Needle-in-Haystack} 
  & \multicolumn{3}{c|}{Uniform}\\
  \hline
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
   & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG
  & \textbf{0.31} $\pm$0.03
  & \textbf{0.72} $\pm$0.02
  & \textbf{0.75} $\pm$0.02
  %%
  & \textbf{0.68} $\pm$0.03
  & \textbf{0.62} $\pm$0.03
  & \textbf{0.65} $\pm$0.03 
  %%
  & \makecell{\textbf{0.44} $\pm$0.03} 
 & \makecell{\textbf{0.52} $\pm$0.02} 
 & \makecell{\textbf{0.58} $\pm$0.02} \\
\hline
  $\TS$ vs $\DEG$  
  & \textbf{0.3} $\pm$0.03
  & \textbf{0.89} $\pm$0.01 
  & \textbf{0.9} $\pm$0.01
  %%
  & \textbf{0.6} $\pm$0.03
  & \textbf{0.52} $\pm$0.03
  & \textbf{0.55} $\pm$0.02 
  %%
 & \makecell{\textbf{0.41} $\pm$0.03} 
 & \makecell{\textbf{0.47} $\pm$0.02} 
 & \makecell{\textbf{0.55} $\pm$0.02} \\ \hline
  $\DG$ vs $\DEG$
  & \textbf{0.63} $\pm$0.03 
  & \textbf{0.6} $\pm$0.02
  & \textbf{0.56} $\pm$0.03
  %%
  & \textbf{0.42} $\pm$0.03
  & \textbf{0.41} $\pm$0.03
  & \textbf{0.39} $\pm$0.02
  %%
   & \makecell{\textbf{0.5} $\pm$0.03} 
 & \makecell{\textbf{0.46} $\pm$0.02} 
 & \makecell{\textbf{0.45} $\pm$0.02} \\ \hline
\end{tabular}
\end{adjustbox}
\caption{\footnotesize {\bf Simultaneous Entry, Market Share}. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. Each cell contains the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
 The time horizon is $T=2000$.}
\label{fig:market_share}
\end{table*}

\begin{table*}[t]
\centering
\begin{adjustbox}{width=\textwidth,center}
\begin{tabular}{|c|c|c|c||c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail}
  & \multicolumn{3}{c|}{Needle-in-Haystack} 
  & \multicolumn{3}{c|}{Uniform}\\
  \hline
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
   & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG
  & 68 (0)  & 560 (8.5)  & 610 (86.5)
  %%
 & 180 (30)  & 380 (0)  & 550 (6.5)
  %%
  &  260 (0)  
  &  780 (676.5)  
  &  880 (897.5) \\
\hline
  $\TS$ vs $\DEG$  
 & 37 (0)  & 430 (0)  & 540 (105)
  %%
 & 150 (10)  & 460 (25)  & 780 (705)
  %%
 & 230 (0)  & 830 (772)  & 980 (1038) \\ \hline
  $\DG$ vs $\DEG$
 & 340 (110)  & 640 (393)  & 670 (425) 
  %%
  & 410 (8.5)  & 760 (666)  & 740 (646)
  %%
  & 530 (101)  & 990 (1058)  & 1000 (1059) \\ \hline
\end{tabular}
\end{adjustbox}
\caption{\footnotesize {\bf Simultaneous Entry, \Eeog}. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. Each cell specifies the ``effective end of game" (\Eeog): the average and the median (in brackets).
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. 
The time horizon is $T=2000$.}
\label{fig:eog}
\end{table*}

\normalsize
Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from  monopoly to ``incumbent" to simultaneous entry duopoly to ``late entrant", we find a stylized inverted-U relationship \asedit{as in Section~\ref{sec:theory-welfare}}. More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DynamicGreedy, \DynamicEpsGreedy and \Thompson. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs, see \cite{DS-arxiv}). \DynamicGreedy is trivially the dominant strategy under monopoly.

@@

@@

\xhdr{Death spiral effect.}
Further, we investigate the ``death spiral" effect mentioned in the Introduction. Restated in terms of our model, the effect is that one firm attracts new customers at a lower rate than the other, and falls behind in terms of performance because the other firm has more customers to learn from, and this gets worse over time until (almost) all new customers go to the other firm. With this intuition in mind, we define  \textit{effective end of game} (\Eeog) for a particular \MRV and realization table, as the last round $t$ such that the agents at this and previous round choose different firms. Indeed, the game, effectively, ends after this round. We interpret low \Eeog as a strong evidence of the ``death spiral" effect. Focusing on the simultaneous entry scenario, we specify the \Eeog values in Table~\ref{fig:eog} (the second line of each cell). We find that the \Eeog values are indeed small:

@@

@@

\begin{figure}
\centering
\includegraphics[scale=0.35]{ec19paper/figures/unif_eq_welfare}
\includegraphics[scale=0.35]{ec19paper/figures/ht_eq_welfare}
\caption{\footnotesize Smoothed welfare plots resulting from equilibrium strategies in the different market structures. Note that welfare at $t = 0$ incorporates the regret incurred during the incumbent and warm start periods. The Thompson Sampling trajectory displays the regret incurred by running Thompson Sampling in isolation on the given instances.}
\label{eq_regret}
\end{figure}

We assume that both firms play their respective equilibrium strategies for the corresponding competition level. As discussed previously, these are:
\begin{OneLiners}
\item \DynamicGreedy in the monopoly,
\item \DynamicGreedy for both firms in simultaneous entry (Finding \ref{find:duopoly}),
\item \Thompson for the incumbent (Finding \ref{find:temp-monopoly}) and \DynamicGreedy for the entrant in first-mover (Finding \ref{find:temp-monopoly-entrant}).
\end{OneLiners}

>>>>

Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from  monopoly to ``incumbent" to simultaneous entry duopoly to ``late entrant", we find a stylized inverted-U relationship \asedit{as in Section~\ref{sec:theory-welfare}}. More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DynamicGreedy, \DynamicEpsGreedy and \Thompson. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs, see \cite{DS-arxiv}). \DynamicGreedy is trivially the dominant strategy under monopoly.

@@

\begin{table*}[t]
\centering
\begin{adjustbox}{width=\textwidth,center}
\begin{tabular}{|c|c|c|c||c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail}
  & \multicolumn{3}{c|}{Needle-in-Haystack}
  & \multicolumn{3}{c|}{Uniform}\\
  \hline
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
   & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG
  & \textbf{0.31} $\pm$0.03
  & \textbf{0.72} $\pm$0.02
  & \textbf{0.75} $\pm$0.02
  %%
  & \textbf{0.68} $\pm$0.03
  & \textbf{0.62} $\pm$0.03
  & \textbf{0.65} $\pm$0.03
  %%
  & \makecell{\textbf{0.44} $\pm$0.03}
 & \makecell{\textbf{0.52} $\pm$0.02}
 & \makecell{\textbf{0.58} $\pm$0.02} \\
\hline
  $\TS$ vs $\DEG$
  & \textbf{0.3} $\pm$0.03
  & \textbf{0.89} $\pm$0.01
  & \textbf{0.9} $\pm$0.01
  %%
  & \textbf{0.6} $\pm$0.03
  & \textbf{0.52} $\pm$0.03
  & \textbf{0.55} $\pm$0.02
  %%
 & \makecell{\textbf{0.41} $\pm$0.03}
 & \makecell{\textbf{0.47} $\pm$0.02}
 & \makecell{\textbf{0.55} $\pm$0.02} \\ \hline
  $\DG$ vs $\DEG$
  & \textbf{0.63} $\pm$0.03
  & \textbf{0.6} $\pm$0.02
  & \textbf{0.56} $\pm$0.03
  %%
  & \textbf{0.42} $\pm$0.03
  & \textbf{0.41} $\pm$0.03
  & \textbf{0.39} $\pm$0.02
  %%
   & \makecell{\textbf{0.5} $\pm$0.03}
 & \makecell{\textbf{0.46} $\pm$0.02}
 & \makecell{\textbf{0.45} $\pm$0.02} \\ \hline
\end{tabular}
\end{adjustbox}
\caption{\footnotesize {\bf Simultaneous Entry, Market Share}. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. Each cell contains the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
 The time horizon is $T=2000$.}
\label{fig:market_share}
\end{table*}

\normalsize

@@

\xhdr{Death spiral effect.}
Further, we investigate the ``death spiral" effect mentioned in the Introduction. Restated in terms of our model, the effect is that one firm attracts new customers at a lower rate than the other, and falls behind in terms of performance because the other firm has more customers to learn from, and this gets worse over time until (almost) all new customers go to the other firm. With this intuition in mind, we define  \textit{effective end of game} (\Eeog) for a particular \MRV and realization table, as the last round $t$ such that the agents at this and previous round choose different firms. Indeed, the game, effectively, ends after this round. We interpret low \Eeog as a strong evidence of the ``death spiral" effect. Focusing on the simultaneous entry scenario, we specify the \Eeog values in Table~\ref{fig:eog}. We find that the \Eeog values are indeed small:

@@

\begin{table*}[h]
\centering
\begin{adjustbox}{width=\textwidth,center}
\begin{tabular}{|c|c|c|c||c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail}
  & \multicolumn{3}{c|}{Needle-in-Haystack}
  & \multicolumn{3}{c|}{Uniform}\\
  \hline
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
   & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG
  & 68 (0)  & 560 (8.5)  & 610 (86.5)
  %%
 & 180 (30)  & 380 (0)  & 550 (6.5)
  %%
  &  260 (0)
  &  780 (676.5)
  &  880 (897.5) \\
\hline
  $\TS$ vs $\DEG$
 & 37 (0)  & 430 (0)  & 540 (105)
  %%
 & 150 (10)  & 460 (25)  & 780 (705)
  %%
 & 230 (0)  & 830 (772)  & 980 (1038) \\ \hline
  $\DG$ vs $\DEG$
 & 340 (110)  & 640 (393)  & 670 (425)
  %%
  & 410 (8.5)  & 760 (666)  & 740 (646)
  %%
  & 530 (101)  & 990 (1058)  & 1000 (1059) \\ \hline
\end{tabular}
\end{adjustbox}
\caption{\footnotesize {\bf Simultaneous Entry, \Eeog}. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. Each cell specifies the ``effective end of game" (\Eeog): the average and the median (in brackets).
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
The time horizon is $T=2000$.}
\label{fig:eog}
\end{table*}

@@

We assume that both firms play their respective equilibrium strategies.
% for the corresponding competition level.
As discussed previously, it is
\DynamicGreedy in the monopoly scenario, and
\DynamicGreedy for both firms in simultaneous entry scenario (Finding \ref{find:duopoly}).
For the first-mover scenario, it is \Thompson for the incumbent (Finding \ref{find:temp-monopoly}) and \DynamicGreedy for the entrant (Finding \ref{find:temp-monopoly-entrant}).

\begin{figure}[bth]
\centering
\includegraphics[scale=0.35]{ec19paper/figures/unif_eq_welfare}
\includegraphics[scale=0.35]{ec19paper/figures/ht_eq_welfare}
\caption{\footnotesize Smoothed welfare plots resulting from equilibrium strategies in the different market structures. Note that welfare at $t = 0$ incorporates the regret incurred during the incumbent and warm start periods. The Thompson Sampling trajectory displays the regret incurred by running Thompson Sampling in isolation on the given instances.}
\label{eq_regret}
\end{figure}
